---
title: |
  | \vspace{-1.3cm} \textbf{Analyzing Voter Registration in the 2016 Presidential Election} \vspace{-0.5cm}
author: Andrew Amore 
date: |
  | \vspace{-0.6cm} \small 2022-11-21
output: pdf_document
urlcolor: blue
fontsize: 11pt
geometry: "left=2cm,right=2cm,top=1.5cm,bottom=1.5cm"
subparagraph: yes
header-includes: |
  \usepackage{titlesec}
  \titlespacing{\title}{0pt}{\parskip}{-\parskip}
  \titlespacing{\section}{0pt}{12pt plus 2pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsection}{0pt}{12pt plus 2pt minus 1pt}{0pt plus 1pt minus 1pt}
  \titlespacing{\subsubsection}{0pt}{12pt plus 2pt minus 1pt}{0pt plus 1pt minus 1pt}
---

\vspace{-1.1cm}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include=FALSE, echo=FALSE, warning=FALSE, 
                      message=FALSE, fig.align = "center")
library(tidyverse)
library(rstan)
library(brms)
library(ggmcmc)
library(ggthemes)
library(ggridges)
library(bayesplot)
library(tidybayes)
library(coda)
library(arm)
library(ggpubr)     # panel plotting

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r, data-load-and-clean}
## CENSUS DATA ##
census = read.table(file="./data/Census2010_long.txt", header = TRUE)

# clean data
## combine NativeHawaiianOrOtherPacificIslanderAlone and SomeOtherRaceAlone to match voter_stats
census = census %>% 
  filter(Race %in% c("NativeHawaiianOrOtherPacificIslanderAlone", "SomeOtherRaceAlone")) %>%
  group_by(Geography, Age, Gender, Hispanic) %>% 
  summarise(freq_sum = sum(Freq)) %>% 
  mutate(Race = "SomeOtherRaceAlone") %>% 
  inner_join(census, by= c("Geography")) %>%
  distinct(Geography, Age.x, Gender.x, Hispanic.x, Race.x, freq_sum, TotalCountyPopulation) %>%
  rename("Age" = "Age.x", "Gender" = "Gender.x", "Hispanic" = "Hispanic.x", "Race" = "Race.x", "Freq" = "freq_sum") %>%
  relocate(Freq, .after = Race) %>%
  union(census %>% filter(!Race %in% c("NativeHawaiianOrOtherPacificIslanderAlone", "SomeOtherRaceAlone"))) %>%
  ungroup()

## VOTER REGISTRATION DATA ##
voter_stats = read.table(file="./data/voter_stats_20161108.txt", header = TRUE, na.strings=c("", " ", NA))
## democrat, libertarian, republican, unaffiliated

# clean data
voter_stats = voter_stats %>% 
  # reformat age to match
  mutate(Age = case_when(
    age == "Age 18 - 25" ~ "18-25",
    age == "Age 26 - 40" ~ "26-40",
    age == "Age 41 - 65" ~ "41-65",
    TRUE ~ "66+")) %>%
  # rename county to match
  rename("Geography" = "county_desc") %>%
  # reformat ethnicity
  mutate(Hispanic = case_when(
    ethnic_code == "NL" ~ "NotHispanic",
    ethnic_code == "HL" ~ "Hispanic",
    TRUE ~ "Undesignated")) %>%
  ## one unmatched field between sources (NativeHawaiianOrOtherPacificIslanderAlone)
  mutate(Race = case_when(
    race_code == "W" ~ "WhiteAlone",
    race_code == "B" ~ "BlackAlone",
    race_code == "O" ~ "SomeOtherRaceAlone",
    race_code == "U" ~ "Undesignated",
    race_code == "I" ~ "AmericanIndianOrAlaskaNativeAlone",
    race_code == "A" ~ "AsianAlone",
    race_code == "M" ~ "TwoOrMoreRaces",
    TRUE ~ "Undesignated")) %>%
  ## one unmatched field (U)
  mutate(Gender = case_when(
    sex_code == "F" ~ "Female",
    sex_code == "M" ~ "Male",
    TRUE ~ "Unknown")) %>% 
  dplyr::select(-age, -ethnic_code, -race_code, -sex_code) %>%
  ## aggregate by county
  group_by(Geography, party_cd, Age, Gender, Hispanic, Race) %>%
  summarise(VoterFreq = sum(total_voters)) %>%
  ungroup()

## PARTY NUMBERS ##
### this probably isn'r relevant as it looks to contain the same information as the
### voter stats file
party_df = read.table(file="./data/party_count.txt", skip = 5, sep = ",", 
                      col.names = c("county", "democrat", "republican", "green", 
                                    "constitution", "libertarian", "unaffiliated",
                                    "white", "black", "american indian", "native hawaiian",
                                    "other", "hispanic", "male", "female", "total"))
## subset to only relevant party counts, pivot and adjust values to match
party_df = party_df %>%
   dplyr::select(county, democrat, republican, libertarian, unaffiliated) %>%
  pivot_longer(!county, names_to = "party", values_to = "party_cnt") %>%
  mutate(party_cd = case_when(
    party == "democrat" ~ "DEM",
    party == "republican" ~ "REP",
    party == "libertarian" ~ "LIB",
    TRUE ~ "UNA"
  )) %>%
  dplyr::select(-party) %>%
  relocate(party_cd, .before=party_cnt)
```

## *Abstract*
*Political candidates are often interested in understanding factors influencing
voter registration rates across different demographic categories and how they 
effect overall registration probability. Understanding these aspects can better 
inform campaign staffs on election related decisions, like where to focus limited
advertising budgets.*

*To conduct an analysis, official voter registration records and census population 
estimates for counties in North Carolina were collected and analyzed using Bayesian
hierarchical modeling strategies to better quantify estimate uncertainty across 
different geographies. The final model indicates...*
<!-- more here once completed... -->

## Analysis

**Introduction**  
Political campaigns analyze historical election data to understand how different
demographic groups register to vote, specifically during presidential election
years when individuals are more likely to vote. This information can inform optimal 
advertising strategies, which can drum up more votes and win elections. Campaigns
are generally interested in assessing registration differences amongst demographic
groups and how registration tendencies vary by county, gender, age and party
affiliation.

**Dataset Overview**  
To address the main questions of interest, two data sources were analyzed. 
Information from the [2010 U.S. Census](https://www.census.gov/data.html) was 
collected from the Federal Census Bureau website and enhanced with official
[2016 voter registration](https://vt.ncsbe.gov/) records from North Carolina. To 
combine the data, demographic field values were standardized, as the coding structure
varies slightly between State and Federal agencies. Field values without a corresponding
match were dropped from the analysis. For example, Census estimates quantify
two genders (male/female) while registration records include a third unknown 
category. Irrelevant registration fields denoting precinct location were also
removed. Metadata information for the combined dataset and a sample observation, 
can be viewed in Table 1.

```{r, show-metadata, include=TRUE, fig.height=3.5}
## ignore U issue and just inner join (drop the U out)
model_df = voter_stats %>% 
  inner_join(census)

# compute features
## factorize
model_df$Age = as.factor(model_df$Age)
model_df$Gender = as.factor(model_df$Gender)
model_df$Hispanic = as.factor(model_df$Hispanic)
model_df$Race = as.factor(model_df$Race)
model_df$PartyCd = as.factor(model_df$party_cd)
model_df = model_df %>% dplyr::select(-party_cd)

## several instances where voter_freq > than CensusFreq...people either travel to polls or 2010 census data doesn't reflect population at 2016 election
model_df$VoterTurnout = model_df$VoterFreq/model_df$TotalCountyPopulation
model_df = model_df %>% filter(Freq > 0)

set.seed(777)
options(scipen=999)
options(digits = 3)
sample = model_df %>% slice_sample(n=1)

tbl_df = data.frame(
    Column.Names = colnames(model_df),
    Column.Description = c(
      "County in North Carolina",
      "Age Demographic Category",
      "Gender Demographic Category",
      "Demographic Indicator of Hispanic Origin",
      "Race Demographic Category",
      "Number of Registered Voters",
      "Total Population Count for Specified Categories",
      "Total County Population",
      "Political Party Affiliation",
      "County Registration Percentage"
    ),
    Sample = t(sample)
)

ggtexttable(tbl_df, rows=NULL, cols = c("Field Name", "Description", "Sample"),
                  theme=ttheme("classic", base_size = 9, padding=unit(c(5,5), "mm"))) %>%
  table_cell_font(row = 2:(nrow(tbl_df) + 1), column = 1, face = "bold", size = 8) %>%
  table_cell_font(row = 2:(nrow(tbl_df) + 1), column = 2, face = "italic", size = 8) %>%
  table_cell_font(row = 2:(nrow(tbl_df) + 1), column = 3, face = "italic", size = 8) %>%
  tab_add_title(text = "Table 1: Metadata Information", size = 9, face = "plain",
                padding=unit(c(1,0), "mm"))
```

**Population Migration**  
The 2010 Census is assumed to represent the voter population during the 2016 
election, however, almost 10% of observations have more registered voters than
the demographic population estimates from the Census. Both examples in Table 2 
have relatively small differences, but the large number of effected observations
warrant additional exploration.

```{r, population-sample, include=TRUE, fig.height=0.9}
set.seed(7324)
options(scipen=999)
options(digits = 3)
sample = model_df %>% 
  filter(VoterFreq > Freq) %>% 
  slice_sample(n=2) %>% 
  dplyr::select(-TotalCountyPopulation, -VoterTurnout)

ggtexttable(sample, rows=NULL,
                  theme=ttheme("classic", base_size = 9, padding=unit(c(3.5,3.5), "mm"))) %>%
  table_cell_bg(row = 2:(nrow(sample) + 1), column=6, linewidth = 1, fill="#FF9999", color = "black") %>%
  table_cell_bg(row = 2:(nrow(sample) + 1), column=7, linewidth = 1, fill="#FF9999", color = "black") %>%
  tab_add_title(text = "Table 2: Bad Data Sample", size = 9, face = "plain",
                padding=unit(c(1,1), "mm"))
```

Figure 1 displays county summaries for the difference between registered voters 
and demographic population. Several counties have large differences, defined as 
100 or more individuals, which may indicate significant population migration.
It's plausible that the 2010 Census incorrectly reflects population densities in
2016, however, the majority of average county differences are less than 50, which
is relatively small. To address this concern, 30 counties were randomly drawn from
the 90 county population using sampling weights inversely proportional to the 
percentage of "incorrect" observations (registered voters > demographic population).
This sub-sampling strategy also reduces computational overhead for an MCMC sampler
in Bayesian models.

```{r, bad-data-boxplot, include=TRUE, fig.height=3.5, fig.cap="Registration Differences by County", fig.align='left'}
# model_df %>% 
#     mutate(flag = case_when (VoterFreq > Freq ~ 1, TRUE ~ 0)) %>%
#   summarize(sum(flag))

## get df for boxplot
sample = model_df %>% 
  filter(VoterFreq > Freq) %>%
  mutate(diff = VoterFreq - Freq)

## remove counties with less than 5 observation to make plot easier to read
sample = sample %>% 
  inner_join(sample %>% group_by(Geography) %>% 
               count() %>% filter(n > 5), keep = FALSE) %>%
  dplyr::select(-n)

# ggplot(sample, aes(x = Geography, y=log(diff))) + 
ggplot(sample, aes(x = reorder(Geography, -log(diff), mean), y=log(diff))) + 
    geom_boxplot(outlier.size = 0.1) + 
  ggtitle("Registration Differences by County") + ylab("Population Difference") +
  xlab("County") +
  theme_minimal() +
  geom_hline(yintercept = log(100), linetype="dashed", color="red", size=0.3) +
  geom_hline(yintercept = log(50), linetype="dashed", color="darkgreen", size=0.3) +
  theme(legend.position="none", 
        plot.title = element_text(hjust = 0.5),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        axis.text = element_text(size = 10),
        axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  scale_y_continuous(breaks = log(c(1, 10, 50, 100, 400)),
                   labels = c(1, 10, 50, 100, 400))
```


```{r, sample-counties-modeling-df}
## use 1-bad_data_percent as sampling weight for county selection
weights = 1 - (model_df %>% group_by(Geography) %>% count() %>% inner_join(model_df) %>% filter(VoterFreq > Freq) %>% group_by(Geography, n) %>% count() %>% arrange(n) %>% mutate(bad_data_percent = nn/n))$bad_data_percent

counties = (model_df %>% group_by(Geography) %>% count() %>% inner_join(model_df) %>% filter(VoterFreq > Freq) %>% group_by(Geography, n) %>% count() %>% arrange(n) %>% mutate(bad_data_percent = nn/n))$Geography

## sample 
set.seed(1235)
samp = sample(counties, 30, replace = FALSE, prob = weights)

## for these voter_freq, set voter_freq = Freq (could also drop these records out)
model_df = model_df  %>%
  mutate(VoterFreq = case_when(
    VoterFreq >= Freq ~ (Freq - as.integer(1)),
    TRUE ~ VoterFreq))

## filter to relevant fields
model_df = model_df %>% 
  filter(Geography %in% samp)
```


**Motivating a Multilevel Model**  
An exploratory data analysis (EDA) was conducted to evaluate modeling decisions,
like where to apply random effects. Figure 1 displays voter registration trends 
for three counties (color coded) and two demographic categories. The left plot fits 
regressions by county & age group, while the right fits identical data by county & 
race. Notice the differences in linear fits between the two plots suggesting
voter behavior is similar across race, but varies significantly across age groups. 
Plots for remaining covariates can be viewed in the Appendix. In addition
to age, party affiliation showed similar variation between categories, but other
covariates displayed consistent trends.

```{r, motivate-the-model, include=TRUE, fig.cap="Voter Registration Behavior", fig.height=4.5}
# ggplot(data=model_df %>% filter(Geography %in% c("WAKE", "MECKLENBURG", "BUNCOMBE")),
p1 = ggplot(data=model_df %>% filter(Geography %in% c("JACKSON", "COLUMBUS", "LEE")),
       aes(x=Freq, y=VoterFreq, col=Geography, group=interaction(Geography, Age))) +
  geom_point(size = 1.2, alpha=.9) + 
  theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5), 
          legend.position="bottom",
          panel.border = element_rect(colour = "black", fill=NA, size=1)) +
  geom_smooth(method = lm, se=FALSE, size=.5, alpha  = .5) +
  # theme(legend.position="bottom", panel.border = element_rect(colour = "black", fill=NA, size=1)) +
  labs(subtitle = "Age") + 
  ylab("Registered Voters") + xlab("Total Demographic Population")
## by race
p2 = ggplot(data=model_df %>% filter(Geography %in% c("JACKSON", "COLUMBUS", "LEE")),
       aes(x=Freq, y=VoterFreq, col=Geography, group=interaction(Geography, Race))) +
  geom_point(size = 1.2, alpha=.9) + 
  theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  geom_smooth(method = lm, se=FALSE, size=.5, alpha  = .5) +
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1), 
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position = "none") +
  labs(subtitle = "Race") + 
  ylab("Registered Voters") + xlab("Total Demographic Population")
plot = ggarrange(p1, p2, ncol=2, nrow=1, common.legend = TRUE, legend="bottom", 
                 widths = c(1.05,0.95))
annotate_figure(plot, top = text_grob("Voter Registration vs. Demographic Population"
                                      , face = "bold", size = 13))
```

\newpage

**Model Specification**  
The modeling framework can be specified as follows. Let $n_{ijk}, p_{ijk}$
denote the total demographic population and probability of voter registration for
voters in county $i$, age group $j$ and remaining covariate indicator combination 
$k$ respectively. We assume the number of voters registering to vote, 
$n_{voters, \{i,j,k\}}$, is distributed as a binomial random variable. 
$$n_{voters, \{i,j,k\}} \sim Binomial(n_{ijk}, p_{ijk})$$
Binomial trials assume replacement, but voters can only register once. This 
simplifying assumption seems reasonable as most demographic categories have
large populations ($n_{ijk}>30$). A more realistic model could assume a 
hypergeometric distribution to account for smaller population sizes, but this 
induces additional computation complexity and was avoided. Several multilevel 
models with binomial distributions were assessed (Table 3). The EDA suggests variation
across counties, age groups and political affiliations, but how does this 
covariance structure affect coefficient estimates?

```{r, random-effect structure, include=TRUE, fig.height=2}
tbl_df = data.frame(
    Fixed.Effects = c(rep("Gender, Hispanic, Race, PartyCd, Age", 3)),
    Random.Intercept = c(rep("Geography", 3)),
    Random.Slope = c("", "Age", "Age, PartyCd")
)

ggtexttable(tbl_df, rows=c("Model I ", "Model II ", "Model III "), 
            cols = c("Fixed Effect Fields", "Random Intercept Fields", "Random Slope Fields"),
                  theme=ttheme("classic", base_size = 9, padding=unit(c(5,5), "mm"))) %>%
  # table_cell_font(row = 2:(nrow(tbl_df) + 1), column = 1, face = "bold", size = 6.5) %>%
  # table_cell_font(row = 2:(nrow(tbl_df) + 1), column = 2, face = "italic", size = 6.5) %>%
  # table_cell_font(row = 2:(nrow(tbl_df) + 1), column = 3, face = "italic", size = 6.5) %>%
  tab_add_title(text = "Table 3: Random Effect Structures", size = 9, face = "plain",
                padding=unit(c(1,0), "mm"))
```


**Model I**  
Let $\mu$ denote a global intercept, $\theta_{i}$ a **county** random effect
intercept for observations in county $i$, $X_k,[ij]$ a vector of demographic indicator
variables (gender, ethnicity, race, party affiliation and age) and $\beta$ the 
corresponding fixed effect estimates.

$$Logit(p_{ijk})) = \mu + \theta_{i} + X_{k,[ij]}\beta$$
<!-- specify priors-->
$$\mu \sim N\left(0,1\right), \ \beta \sim N\left(0, 1\right), \ \theta_{i} \sim N\left(0, \sigma\right)$$
$$\sigma \sim HalfCauchy\left(0, \frac{1}{2}\right)$$

```{r, fit-model-1}
m1 = brm(data = model_df, family = "binomial",
      VoterFreq | trials(Freq) ~ 1 + Gender + Hispanic + Race + PartyCd + Age + (1 | Geography),
      prior = c(prior(normal(0, 1), class = "Intercept"),
                prior(normal(0, 1), class = "b"),  ## class b denotes population (fixed) effects
                prior(cauchy(0, 0.5), class = "sd")  ## group effects
                ),
      file = "models/m1.rds",
      iter = 2500, 
      warmup = 500,
      chains = 4)
summary(m1)
```

**Model II**  
Model II is an extension of I with a new random effect for **Age**. Let $a_{ij,[k]}$ 
denote the age category, $j$, for observations in county $i$ and $\Gamma_{ij}$ the random effect
slope. Other variables remain the same.

$$Logit(p_{ijk})) = \mu +  \theta_{i} + a_{ij, [k]}\Gamma_{ij} + X_{k,[ij]}\beta$$
<!-- specify priors -->
$$\Gamma_{ij} \sim N\left(0, \gamma\right), \ \gamma \sim HalfCauchy\left(0, \frac{1}{2}\right)$$

```{r, fit-model-2}
m2 = brm(data = model_df, family = "binomial",
      VoterFreq | trials(Freq) ~ 1 + Gender + Hispanic + Race + PartyCd + Age + (1 + Age | Geography),
      prior = c(prior(normal(0, 1), class = "Intercept"),
                prior(normal(0, 1), class = "b"),  ## class b denotes population effects
                prior(cauchy(0, 0.5), class = "sd")  ## group effects
                ),
      file = "models/m2_logistic_regression_party_cd.rds",
      iter = 2500, 
      warmup = 500,
      chains = 4)

m2
```

**Model III**  
An extension of II with an additional random effect for **party affiliation**.
Let $q_{il,[jk]}$ denote the party, $l$, for observations in county $i$ and 
$\Omega_{il}$ the random effect slope. Other variables remain identical.


$$Logit(p_{ijlk})) = \theta_{i} + a_{ij, [k]}\Gamma_{ij} + q_{il,[jk]}\Omega_{il} + X_{k,[i,j]}\beta$$
<!-- specify priors -->
$$\Omega_{il} \sim N\left(0, \omega\right), \ \omega \sim HalfCauchy\left(0, \frac{1}{2}\right)$$

```{r, fit-model}
m3 = brm(data = model_df, family = "binomial",
      VoterFreq | trials(Freq) ~ 1 + Gender + Hispanic + Race + PartyCd + Age + (1 + Age + PartyCd | Geography),
      prior = c(prior(normal(0, 1), class = "Intercept"),
                prior(normal(0, 1), class = "b"),  ## class b denotes population effects
                prior(cauchy(0, 0.5), class = "sd")  ## group effects
                ),
      file = "models/m3_logistic_regression_age_party_cd_slope.rds",
      iter = 3000, 
      warmup = 500,
      chains = 4)
m3
```


**Assessing Model Fit & Comparing Models**




```{r}

```


**Analysis**  

<!-- How did different demographic subgroups register to vote in the 2016 elections? For example, how did the registration rate for males compare to the registration rate for females after controlling for other potential predictors? -->

```{r, demographic-subgroup-analysis}
m1_fxed = data.frame(fixef(m1))
m1_fxed$model = "Model II"
m1_fxed$parameter = rownames(m1_fxed)
m1_fxed = m1_fxed[2:nrow(m1_fxed),]
rownames(m1_fxed) = NULL

m2_fxed = data.frame(fixef(m2))
m2_fxed$model = "Model II"
m2_fxed$parameter = rownames(m2_fxed)
m2_fxed = m2_fxed[2:nrow(m2_fxed),]
rownames(m2_fxed) = NULL

m3_fxed = data.frame(fixef(m3))
m3_fxed$model = "Model III"
m3_fxed$parameter = rownames(m3_fxed)
m3_fxed = m3_fxed[2:nrow(m3_fxed),]
rownames(m3_fxed) = NULL

tst = rbind(m1_fxed, m2_fxed, m3_fxed)

ggplot(tst, aes(x=Estimate, y=parameter,xmin = Q2.5, xmax = Q97.5, color=model)) + 
  geom_pointinterval() +
  ggtitle("Population Estimates") +
  ylab("Fixed Effect") + xlab("Coefficient Estimate") + 
  theme(plot.title = element_text(hjust = 0.5)) + geom_vline(xintercept = 0)
```



<!-- Did the overall probability or odds of registering to vote differ by county in 2016? Which counties differ the most from other counties? -->

<!-- How did the registration rates differ between females and males for the different party affiliations? -->

<!-- How did the registration rates differ between age groups for the different party affiliations? -->


**Predictive Performance**  

```{r}
## nice to have...
# p = predict(m1) %>% 
#   as_tibble() %>% bind_cols(model_df)

## for each model how many estimates fall inside the CR
```



**Conclusion**  


\newpage

## Appendix

```{r, additional-covariate-plots, include=TRUE}
ggplot(data=model_df %>% filter(Geography %in% c("JACKSON", "COLUMBUS", "LEE")),
       aes(x=Freq, y=VoterFreq, col=Geography, group=interaction(Geography, Gender))) +
  geom_point(size = 1.2, alpha=.9) + 
  theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  geom_smooth(method = lm, se=FALSE, size=.5, alpha  = .5) +
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
        legend.position = "none") +
  labs(subtitle = "Color by Geography & Fit by Gender") + 
  ylab("Registered Voters") + xlab("Total Demographic Population")

ggplot(data=model_df %>% filter(Geography %in% c("JACKSON", "COLUMBUS", "LEE")),
       aes(x=Freq, y=VoterFreq, col=Geography, group=interaction(Geography, Hispanic))) +
  geom_point(size = 1.2, alpha=.9) + 
  theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  geom_smooth(method = lm, se=FALSE, size=.5, alpha  = .5) +
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1), legend.position = "none") +
  labs(subtitle = "Color by Geography & Fit by Hispanic") + 
  ylab("Registered Voters") + xlab("Total Demographic Population")

ggplot(data=model_df %>% filter(Geography %in% c("JACKSON", "COLUMBUS", "LEE")),
       aes(x=Freq, y=VoterFreq, col=Geography, group=interaction(Geography, PartyCd))) +
  geom_point(size = 1.2, alpha=.9) + 
  theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) +
  geom_smooth(method = lm, se=FALSE, size=.5, alpha  = .5) +
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1), legend.position = "none") +
  labs(subtitle = "Color by Geography & Fit by Political Party") + 
  ylab("Registered Voters") + xlab("Total Demographic Population")
```


**Trace Plots**  

```{r, trace-plots, include=TRUE}
m1df = ggs(m1)
# get the same plots without bayesplot...have to filter out the warmup!
m1df$Chain = factor(m1df$Chain)
# caterpillar trace plots
## only pick some paremeters for plotting
ggplot(filter(m1df, Parameter %in% c("b_Age66P", "b_GenderMale", "b_HispanicNotHispanic")),
       aes(x   = Iteration,
           y   = value,
           col = as.factor(Chain)))+
  theme_minimal() +
  geom_line() +
  geom_vline(xintercept = 500) + ## add vertical line at the default warm-up value 1000
  facet_grid(Parameter ~ . , scale  = 'free_y', switch = 'y') +
  labs(title = "Model I: Trace Plots", col   = "Chains")
```




